Typing a linear \picalc

### 2020-03-31 paper/paper.tex
Add a unit type, use definition environment

\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{todonotes}
\usepackage{enumitem}
\setlist{parsep=0pt,listparindent=\parindent}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage{amsbsy}
\usepackage{mathtools}
\usepackage[many]{tcolorbox}
\bibliographystyle{plainurl}

@@

@@

% Constructors
\newcommand{\PO}{\constr{\mathbf{0}}}
\newcommand{\comp}[2]{#1 \; \constr{\parallel} \; #2}
\newcommand{\new}{\constr{\boldsymbol{\nu}} \;}
\newcommand{\send}[2]{#1 \; \constr{\langle} \; #2 \;\constr{\rangle} \;}
\newcommand{\recv}[1]{#1 \; \constr{\mathbb{()}} \;}
\newcommand{\suc}{\constr{\scriptstyle 1+}}
\newcommand{\base}[1]{\constr{B[} \; #1 \; \constr{]}}
\newcommand{\channel}[2]{\constr{C[} \; #1 \; \constr{\propto} \; #2 \; \constr{]}}
\newcommand{\comma}{\; \constr{,} \;}

@@

\begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\type{\N} : \Set}

@@

Type constructors as the names for the typing rules.
Double rule for type-level definitions.
We omit universe levels for brevity.
$\type{TYPES}$ are in blue and uppercased (and their indices are subscripts), $\constr{constructors}$ are in olive, $\field{fields}$ in red and variables and functions in black.
Some constructors are shared across types. They can however always be disambiguated through the type of the goal.

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Var_n : \Set}

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Process_n : \Set}
\end{mathpar}

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {P \eqeq Q : \Set}

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {\Rec : \Set}

@@

  \datatype
  {P \, Q : \Process_n \\ r : \Rec}
  {P \eq{r} Q : \Set}

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Channel_n : \Set}

@@

  \datatype
  {i : \Channel_n \\ P \; Q : \Process_n}
  {P \reduce{i} Q : \Set}

@@

Together, these constraints give raise to a \emph{partial commutative monoid}.
In Figure \ref{fig:multiplicities} we present an algebra $\Quantifier_C$ on the carrier set $C$ in the form of a terniary relation such that it forms a partial commutative monoid.

\todo{SHOULD ALL THESE NOT BE DEFINITIONS?}

\begin{figure}[h]
\begin{equation}
\begin{aligned}
  &\zero                  &:{} &                      &        & C \\
  &\one                   &:{} &                      &        & C \\
  &\op{\_}{\_}{\_}        &:{} &                      &        & C \to C \to C \to \Set \\
  &\field{\cdot-compute}  &:{} &\forall y z           & \to \; & \type{DEC} \; (\type{\exists} x . \; (\op{x}{y}{z})) \\
  &\field{\cdot-unique}   &:{} &\forall \{x x' y z\}  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
  &\field{\cdot-unique^l} &:{} &\forall \{x y y' z\}  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
  &\field{\cdot-id^l}     &:{} &\forall x             & \to \; & \op{x}{\zero}{x} \\
  &\field{\cdot-comm}     &:{} &\forall \{x y z\}     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
  &\field{\cdot-assoc}    &:{} &\forall \{x y z u v\} & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
\end{aligned}
\end{equation}
\caption{Quantifier algebra $\Quantifier_C$ algebra on a partial commutative monoid.}
\label{fig:multiplicities}
\end{figure}

To allow for several multiplicity algebras to coexist on the same type system, we index the carriers of the algebras by an index $\Idx$ (see Figure \ref{fig:indexed-multiplicities}).
Each index $i$ has a corresponding carrier type $\field{CARRIER}_i$ such that $\field{CARRIER}_i$ satisfies the algebra $\Quantifier_{\field{CARRIER}_i}$.
We require the indexing type to be inhabited -- necessary to prove type preservation on $\new \PO_{\suc n} \eqeq \PO_n$.
\todo{lift op to $^2$}

\begin{figure}[h]
\begin{equation}
\begin{split}
  &\field{IDX}          &:{} &\Set \\
  &\field{\exists IDX}  &:{} &\field{IDX} \\
  &\field{CARRIER}      &:{} &\field{IDX} \to \Set \\
  &\field{QUANTIFIERS}  &:{} &(i : \field{IDX}) \to \Quantifier_{\field{CARRIER}_i}
\end{split}
\end{equation}
\caption{Indexed set of partial commutative monoids.}
\label{fig:indexed-multiplicities}
\end{figure}

@@

Figure \ref{fig:prectx} represents base types $\base{n}$ using a natural number $n$ that can be reified into a type by the host language.
Using natural numbers instead of type universes is convenient: there is no need to litter type types with universe polymorphism, and it provides an additional proof of inhabitance, needed for the $\new \PO_{\suc n} \eqeq \PO_n$ rule in subject congruence.

@@

Types are then listed in a length-indexed \emph{pre-context} $\PreCtx_n$.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {\Type : \Set}

@@

\begin{mathpar}
  \datatype
      {n : \N}
      {\PreCtx_n : \Set}

@@

      \inferrule
          {\gamma : \PreCtx_n \\ t : \Type}
          {\gamma \comma t : \PreCtx_{\suc n}}
\end{mathpar}
\caption{Types and length-indexed typing contexts.}
\label{fig:prectx}
\end{figure}

To allow different multiplicity algebras to coexist we first define a length-indexed context of carrier indices $\Idxs_n$ (see \ref{fig:ctx}).
On top of it, we define a context of multiplicities $\Ctx_{idxs}$.
We lift the algebra on multiplicities to multiplicity contexts by constraining the algebra on multiplicity contexts to act on  contexts defined on the same underlying context of carrier indices.
We use two multiplicities over a common index to represent multiplicities for both the input and the output capabilities ($\Carrier_{idx}^2$).

\begin{figure}[h]
\begin{mathpar}
  \datatype
      {n : \N}
      {\Idxs_n : \Set}

@@

  \datatype
      {idxs : \Idxs_n}
      {\Ctx_{idxs} : \Set}

@@

  \inferrule
      {\Gamma : \Ctx_{idxs} \\ x : \Carrier_{idx} ^2}
      {\Gamma \comma x : \Ctx_{idxs \comma idx}}
\end{mathpar}
\caption{Length-indexed context of carrier indices with a context of multiplicities on top.}
\label{fig:ctx}
\end{figure}

@@

An alternative is to keep track in the typing judgment of the way in which the context has been transformed.
One way of doing so is by mentioning both the \emph{input contexts} and the \emph{output contexts} over which the processes are typed.
Parallel composition then uses the output context of the left hand process as the input context of the right hand.
As an additional benefit, the associativity of parallel composition becomes trivial to prove.
\S \ref{typing-judgments} introduces the typing rules for such a typing judgment.

In the \picalc{} both input and output contain variable references.
Both these actions need to modify their context, and thus using variable references that mention both their input and output contexts seems relevant.
\S \ref{variable-references} introduces the typing rules for such variable references.

@@

\subsection{Variable references} \label{variable-references}

Akin to the typing rules for processes, the rules for variable references presented in Figure \ref{fig:variable-references} are indexed over the typing context, the input multiplicity context, and the output multiplicity context.
The base case $\constr{zero}$ splits the last multiplicity in the input context into $x$ and the last multiplicity in the output context, while also ensuring that $t$ is the last type in the typing context.
The inductive case $\constr{suc}$ appends the same multiplicity to both the input context and the output context, while also appending an extra type to the typing context.

@@

\todo{lift op to $^2$}

\begin{figure}[h]
\begin{adjustwidth}{-1cm}{-1cm}
\centering
\begin{mathpar}
  \mprset{sep=0.5em}

  \datatype{
    \gamma : \PreCtx_n \\
    \stacked{
      idxs : \Idxs_n \\\\
      \Gamma : \Ctx_{idxs}} \\
    i : \Var_n \\
    t : \Type \\
    \stacked{
      idx : \Idx \\\\
      x : \Carrier_{idx}^2} \\
    \Delta : \Ctx_{idxs}}
  {\contains{\gamma}{\Gamma}{i}{t}{x}{\Delta} : \Set}

  \inferrule
      {\gamma : \PreCtx_n \\
        \Gamma : \Ctx_{idxs}\\
        t : \Type \\
        x \; y \; z : \Carrier_{idx}^2\\
        \op{x}{y}{z}}
  {\constr{zero} : \contains{\gamma \comma t}{\Gamma \comma x}{zero}{t}{y}{\Gamma \comma z}}

  \inferrule
      {\gamma : \PreCtx_n \\
        \Gamma : \Ctx_{idxs}\\
        t \; t' \; : \Type \\
        x : \Carrier_{idx}^2 \\
        x' : \Carrier_{idx'}^2 \\
        \Delta : \Ctx_{idxs} \\\\
    loc_i : \contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}}
  {\constr{suc} \; loc_i : \contains{\gamma \comma t'}{\Gamma \comma x'}{\suc i}{t}{x}{\Delta \comma x'}}
\end{mathpar}
\end{adjustwidth}
\caption{Variable references as proofs on the typing context and input and output multiplicity contexts.}
\label{fig:variable-references}
\end{figure}

\todo{example}

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype{
    \gamma : \PreCtx_n \\
    \stacked{
      idxs : \Idxs_n \\\\
      \Gamma : \Ctx_{idxs}} \\
    P : \Process_n \\
    \Delta : \Ctx_{idxs}}
  {\types{\gamma}{\Gamma}{P}{\Delta} : \Set}

@@

  \inferrule
  {l : \types{\gamma}{\Gamma}{P}{\Delta} \\\\
   r : \types{\gamma}{\Delta}{Q}{\Xi}}
  {\constr{comp} \; l \; r : \types{\gamma}{\Gamma}{\comp{P}{Q}}{\Xi}}
\end{mathpar}
\caption{This is...}
\end{figure}

>>>>

\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{todonotes}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage{amsbsy}
\usepackage{mathtools}
\usepackage[many]{tcolorbox}
\bibliographystyle{plainurl}

% Mathbb doesn't support digits
\usepackage{bbm}

@@

% Do not use italics for definitions
\theoremstyle{definition}
\newtheorem{nidefinition}[theorem]{Definition}

% No line numbers
\nolinenumbers

@@

% Constructors
\newcommand{\PO}{\constr{\mathbf{0}}}
\newcommand{\comp}[2]{#1 \; \constr{\parallel} \; #2}
\newcommand{\new}{\constr{\boldsymbol{\nu}} \;}
\newcommand{\send}[2]{#1 \; \constr{\langle} \; #2 \;\constr{\rangle} \;}
\newcommand{\recv}[1]{#1 \; \constr{\mathbb{()}} \;}
\newcommand{\suc}{\constr{\scriptstyle 1+}}
\newcommand{\unit}{\constr{\mathbbm{1}}}
\newcommand{\base}[1]{\constr{B[} \; #1 \; \constr{]}}
\newcommand{\channel}[2]{\constr{C[} \; #1 \; \constr{\propto} \; #2 \; \constr{]}}
\newcommand{\comma}{\; \constr{,} \;}

@@

\begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\type{\N} : \Set}
    \; \textsc{Nat}

@@

Type constructors as the names for the typing rules.
Datatype constructor rules might have additional index-free synonyms for ease of reference.
Double rule for type-level definitions.
We omit universe levels for brevity.
$\type{TYPES}$ are in blue and uppercased (and their indices are subscripts), $\constr{constructors}$ are in olive, $\field{fields}$ in red and variables and functions in black.
Some constructors are shared across types. They can however always be disambiguated through the type of the goal.

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Var_n : \Set}
  \; \textsc{Var}

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Process_n : \Set}
  \; \textsc{Process}
\end{mathpar}

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {P \eqeq Q : \Set}
  \; \textsc{StructCong}

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {\Rec : \Set}
  \; \textsc{Rec}

@@

  \datatype
  {P \, Q : \Process_n \\ r : \Rec}
  {P \eq{r} Q : \Set}
  \; \textsc{Equals}

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Channel_n : \Set}
  \; \textsc{Channel}

@@

  \datatype
  {i : \Channel_n \\ P \; Q : \Process_n}
  {P \reduce{i} Q : \Set}
  \; \textsc{Reduces}

@@

\begin{nidefinition}
  Together, the aforementioned constraints give raise to a \emph{partial commutative monoid}.
  We deal with partiality by modeling the monoidal operation as a terniary relation.
  An algebra on the carrier set $C$ forms a $\Quantifier_C$ iff it satisfies the fields in Figure \ref{fig:multiplicities}.

  \begin{figure}[h]
  \begin{equation}
  \begin{aligned}
    &\zero                  &:{} &                      &        & C \\
    &\one                   &:{} &                      &        & C \\
    &\op{\_}{\_}{\_}        &:{} &                      &        & C \to C \to C \to \Set \\
    &\field{\cdot-compute}  &:{} &\forall y z           & \to \; & \type{DEC} \; (\type{\exists} x . \; (\op{x}{y}{z})) \\
    &\field{\cdot-unique}   &:{} &\forall \{x x' y z\}  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\field{\cdot-unique^l} &:{} &\forall \{x y y' z\}  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\field{\cdot-id^l}     &:{} &\forall x             & \to \; & \op{x}{\zero}{x} \\
    &\field{\cdot-comm}     &:{} &\forall \{x y z\}     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\field{\cdot-assoc}    &:{} &\forall \{x y z u v\} & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation}
  \caption{Quantifier algebra $\Quantifier_C$ algebra on a partial commutative monoid.}
  \label{fig:multiplicities}
  \end{figure}
\end{nidefinition}

\begin{nidefinition}
  An indexed set of partial commutative monoids is a set of indices such that for every index there exists a carrier $C$ with a $\Quantifier_C$ algebra defined on it.
  This allows for several algebras to coexist in the same type system.
  Figure \ref{fig:indexed-multiplicities} captures these requirements: each index $i$ has a corresponding carrier type $\field{CARRIER}_i$ such that $\field{CARRIER}_i$ satisfies the algebra $\Quantifier_{\field{CARRIER}_i}$.
  We require the index type to be inhabited -- necessary to prove type preservation on $\new \PO_{\suc n} \eqeq \PO_n$.
  \todo{lift op to $^2$}

  \begin{figure}[h]
  \begin{equation}
  \begin{split}
    &\field{IDX}          &:{} &\Set \\
    &\field{\exists IDX}  &:{} &\field{IDX} \\
    &\field{CARRIER}      &:{} &\field{IDX} \to \Set \\
    &\field{QUANTIFIERS}  &:{} &(i : \field{IDX}) \to \Quantifier_{\field{CARRIER}_i}
  \end{split}
  \end{equation}
  \caption{Indexed set of partial commutative monoids.}
  \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}

@@

\begin{nidefinition}
  Figure \ref{fig:prectx} defines types as either unit types, base types or channel types.
  Unit types $\unit$ provide a proof of inhabitance for types, needed for the $\new \PO_{\suc n} \eqeq \PO_n$ rule in subject congruence.

  Base types $\base{n}$ use a natural number $n$ that can be reified into a type by the host language.
  Using natural numbers instead of type universes is convenient: there is no need to litter type types with universe polymorphism.

@@

  Types form length-indexed \emph{pre-contexts} $\PreCtx_n$.

  \begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\Type : \Set}
    \; \textsc{Type}

    \inferrule
    { }
    {\unit : \Type}

@@

  \begin{mathpar}
    \datatype
    {n : \N}
    {\PreCtx_n : \Set}
    \; \textsc{PreCtx}

@@

        \inferrule
            {\gamma : \PreCtx_n \\ t : \Type}
            {\gamma \comma t : \PreCtx_{\suc n}}
  \end{mathpar}
  \caption{Types and length-indexed typing contexts.}
  \label{fig:prectx}
  \end{figure}
\end{nidefinition}

\begin{nidefinition}
  To allow different multiplicity algebras to coexist we first define a length-indexed context of carrier indices $\Idxs_n$ (see \ref{fig:ctx}).
  On top of it, we define a context of multiplicities $\Ctx_{idxs}$.
  We lift the algebra on multiplicities to multiplicity contexts by constraining the algebra on multiplicity contexts to act on  contexts defined on the same underlying context of carrier indices.
  We use two multiplicities over a common index to represent multiplicities for both the input and the output capabilities ($\Carrier_{idx}^2$).

  \begin{figure}[h]
  \begin{mathpar}
    \datatype
    {n : \N}
    {\Idxs_n : \Set}
    \; \textsc{Idxs}

@@

    \datatype
    {idxs : \Idxs_n}
    {\Ctx_{idxs} : \Set}
    \; \textsc{Ctx}

@@

    \inferrule
        {\Gamma : \Ctx_{idxs} \\ x : \Carrier_{idx} ^2}
        {\Gamma \comma x : \Ctx_{idxs \comma idx}}
  \end{mathpar}
  \caption{Length-indexed context of carrier indices with a context of multiplicities on top.}
  \label{fig:ctx}
  \end{figure}
\end{nidefinition}

@@

An alternative is to keep track in the typing judgment of the way in which the context has been transformed.
One way of doing so is by mentioning both the \emph{input contexts} and the \emph{output contexts} over which the processes are typed.
Parallel composition then uses the output context of the left hand process as the input context of the right hand.
As an additional benefit, the associativity of parallel composition becomes trivial to prove.

In the \picalc{} both input and output contain variable references.
Both these actions need to modify their context, and thus using variable references that mention both their input and output contexts seems relevant.

@@

\begin{nidefinition}
  Akin to the typing rules for processes, the rules for variable references presented in Figure \ref{fig:variable-references} are indexed over the typing context, the input multiplicity context, and the output multiplicity context.
  The base case $\constr{zero}$ splits the last multiplicity in the input context into $x$ and the last multiplicity in the output context, while also ensuring that $t$ is the last type in the typing context.
  The inductive case $\constr{suc}$ appends the same multiplicity to both the input context and the output context, while also appending an extra type to the typing context.

  \todo{example}
  \todo{lift op to $^2$}

  \begin{figure}[h]
  \begin{mathpar}
    \mprset{sep=0.5em}

    \datatype{
      \gamma : \PreCtx_n \\
      \stacked{
        idxs : \Idxs_n \\\\
        \Gamma : \Ctx_{idxs}} \\
      i : \Var_n \\
      t : \Type \\
      \stacked{
        idx : \Idx \\\\
        x : \Carrier_{idx}^2} \\
      \Delta : \Ctx_{idxs}}
    {\contains{\gamma}{\Gamma}{i}{t}{x}{\Delta} : \Set}
    \; \textsc{VarRef}

    \inferrule
        {t : \Type \\
          x \; y \; z : \Carrier_{idx}^2\\
          \op{x}{y}{z}}
    {\constr{zero} : \contains{\gamma \comma t}{\Gamma \comma x}{zero}{t}{y}{\Gamma \comma z}}

    \inferrule
        {t \; t' \; : \Type \\
          x : \Carrier_{idx}^2 \\
          x' : \Carrier_{idx'}^2 \\\\
      loc_i : \contains{\gamma}{\Gamma}{i}{t}{x}{\Delta}}
    {\constr{suc} \; loc_i : \contains{\gamma \comma t'}{\Gamma \comma x'}{\suc i}{t}{x}{\Delta \comma x'}}
  \end{mathpar}
  \caption{Variable references as proofs on the typing context and input and output multiplicity contexts.}
  \label{fig:variable-references}
  \end{figure}
\end{nidefinition}

@@

@@

\begin{nidefinition}
  \begin{figure}[h]
  \begin{mathpar}
    \datatype{
      \gamma : \PreCtx_n \\
      \stacked{
        idxs : \Idxs_n \\\\
        \Gamma : \Ctx_{idxs}} \\
      P : \Process_n \\
      \Delta : \Ctx_{idxs}}
    {\types{\gamma}{\Gamma}{P}{\Delta} : \Set}
    \; \textsc{Types}

@@

    \inferrule
    {l : \types{\gamma}{\Gamma}{P}{\Delta} \\\\
     r : \types{\gamma}{\Delta}{Q}{\Xi}}
    {\constr{comp} \; l \; r : \types{\gamma}{\Gamma}{\comp{P}{Q}}{\Xi}}
  \end{mathpar}
  \caption{This is...}
  \end{figure}
\end{nidefinition}
